---
title: "Final_Project_NBA"
author: "Shengbo Wang"
date: "2022-12-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE
)
library(tidyr)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(knitr)
library(stringr)
library(kableExtra)
library(lme4)
library(corrplot)
library(PerformanceAnalytics)
library(latexpdf)
```

##Abstract

##Introduction

I knew Yao Ming when I was very young. He is the most famous Chinese player in NBA in China till today. Yao Ming has also brought the game of basketball to schools. I also started watching NBA games because of Yao Ming. Now Basketball is one of the most popular sports in the world, NBA has the most high quality of players. Some of the players are over 34 millions per year. Apart from the players' popularity and Internet traffic, I want to know what makes more money per year by looking at their statistic in the game. There are five positions in Basketball, each position will have different ability to help their team to win. And each teams will have different tactics like the Golden State Warriors are famous the three points, so they might hire better three-point shooter in higher salary. So, I decided make a multilevel model to find the influence of fixed effects.

##Methods


###Data Preprocessing
The data set come from (https://www.kaggle.com/datasets/drgilermo/nba-players-stats), which has the Season stats from many different aspects, such ass Three point field goals, two points field goals, win shares, etc. And I found the (https://www.kaggle.com/datasets/koki25ando/salary) which has the salary of NBA players during the season 2017-2018, so just filter the year in the Season stats in 2016 because the salary is based on the players' last season performance, and inner join with the salary data, finally I got 251 observations in 25 variables. There are five positions in Basketball, each position will have different ability to help their team to win. And each teams will have different tactics like the Golden State Warriors are famous the three points. I decided make a multilevel model to find the influence of fixed effects.


| column names    | explanation                                         |
| :--:            | :-----                                              |
| Player          | Name of the player                                  |
| Tm              | Name of the Team                                    |
| Pos             | Position of player                                  |
| Age             | Age of player                                       |
| G               | Game played                                         |
| FTr             | Free Throw Rate                                     |
| WSperM          | Win Shares Per 48 Minutes                           |
| FG              | Field Goals                                         |
| FGr             | Field Goal Percentage                               |
| ThreeP          | 3-Point Field Goals                                 |
| ThreePr         | 3-Point Field Goal Percentage                       |  
| TwoP            | 2-Point Field Goals                                 |
| TwoPr           | 2-Point Field Goals Percentage                      |
| EffectiveP      | Effective Field Goal Percentage                     |
| FT              | Free Throws                                         |
| FTp             | Free Throw Percentage                               |
| TRB             | Total Rebounds                                      |
| AST             | Assists                                             |
| STL             | Yards gained from scrimmage                         |
| BLK             | Blocks                                              |
| TOV             | Turnovers                                           |
| PF              | Personal Fouls                                      |
| PTS             | Points                                              |
| Salary          | Annual salary of player (dollars)                   |

```{r}
Salary <- read_csv("NBA_season1718_salary.csv") %>% 
  select(Player,Tm,Salary=salary)
Seasons_Stats <- read_csv("Seasons_Stats.csv") %>%
  filter(Year==2016) %>%
  select(Year,Player,Tm,Pos,Age,G,
         FTr,WSperM="WS/48",FG,FGr="FG%",
         ThreeP="3P",ThreePr="3P%",TwoP="2P",TwoPr="2P%",
         EffectiveP="eFG%",FT,FTp="FT%",TRB,AST,STL,BLK,
         TOV,PF,PTS)

NBA <- inner_join(Seasons_Stats,Salary)
NBA <- data.frame(NBA) %>% na.omit()

set.seed(0611)
sample_team <- sample(unique(NBA$Tm), 10, replace = FALSE)

```

```{r}
ggplot(subset(NBA,Tm %in% sample_team),aes(Salary))+
  geom_density(col="red")+
  facet_wrap(~Tm)

ggplot(NBA,aes(Salary))+
  geom_density(col="blue")+
  facet_wrap(~Pos)


```

##EDA

```{r, fig.show="hold", out.width="50%",fig.cap=paste("Realationship between Gameplayed and salaries")}
ggplot(subset(NBA,Tm %in% sample_team)) + 
  aes(x = log(G + 1), y = log(Salary)) + 
  geom_point(aes(color = factor(Tm)), size = .6) + 
  geom_smooth(aes(color = factor(Tm)), method = "lm", se = FALSE, formula = 'y ~ x')

ggplot(NBA) + 
  aes(x = log(G + 1), y = log(Salary)) + 
  geom_point(aes(color = factor(Pos)), size = .6) + 
  geom_smooth(aes(color = factor(Pos)), method = "lm", se = FALSE, formula = 'y ~ x')
```

```{r, fig.show="hold", out.width="50%",fig.cap=paste("Realationship between Points made and salaries")}

ggplot(subset(NBA,Tm %in% sample_team)) + 
  aes(x = log(PTS + 1), y = log(Salary)) + 
  geom_point(aes(color = factor(Tm)), size = .6) + 
  geom_smooth(aes(color = factor(Tm)), method = "lm", se = FALSE, formula = 'y ~ x')

ggplot(NBA) + 
  aes(x = log(PTS + 1), y = log(Salary)) + 
  geom_point(aes(color = factor(Pos)), size = .6) + 
  geom_smooth(aes(color = factor(Pos)), method = "lm", se = FALSE, formula = 'y ~ x')
```

```{r, fig.show="hold", out.width="50%",fig.cap=paste("Realationship between Win share per 84 mins and salaries")}
ggplot(subset(NBA,Tm %in% sample_team)) + 
  aes(x = WSperM, y = log(Salary)) + 
  geom_point(aes(color = factor(Tm)), size = .6) + 
  geom_smooth(aes(color = factor(Tm)), method = "lm", se = FALSE, formula = 'y ~ x')

ggplot(NBA) + 
  aes(x = WSperM, y = log(Salary)) + 
  geom_point(aes(color = factor(Pos)), size = .6) + 
  geom_smooth(aes(color = factor(Pos)), method = "lm", se = FALSE, formula = 'y ~ x')
```

```{r, fig.show="hold", out.width="50%",fig.cap=paste("Realationship between Fiedl goals per 84 mins and salaries")}
ggplot(subset(NBA,Tm %in% sample_team)) + 
  aes(x = FG, y = log(Salary)) + 
  geom_point(aes(color = factor(Tm)), size = .6) + 
  geom_smooth(aes(color = factor(Tm)), method = "lm", se = FALSE, formula = 'y ~ x')

ggplot(NBA) + 
  aes(x = FG, y = log(Salary)) + 
  geom_point(aes(color = factor(Pos)), size = .6) + 
  geom_smooth(aes(color = factor(Pos)), method = "lm", se = FALSE, formula = 'y ~ x')

```
```{r, fig.show="hold", out.width="50%",fig.cap=paste("Realationship between Totoal rebound per 84 mins and salaries")}
ggplot(subset(NBA,Tm %in% sample_team)) + 
  aes(x = TRB, y = log(Salary)) + 
  geom_point(aes(color = factor(Tm)), size = .6) + 
  geom_smooth(aes(color = factor(Tm)), method = "lm", se = FALSE, formula = 'y ~ x')

ggplot(NBA) + 
  aes(x = TRB, y = log(Salary)) + 
  geom_point(aes(color = factor(Pos)), size = .6) + 
  geom_smooth(aes(color = factor(Pos)), method = "lm", se = FALSE, formula = 'y ~ x')
```


#Model fitting

```{r, fig.height=5, fig.width=7, fig.cap="Correlation Matrix"}
NBA_Subset <- NBA[,c(5:20)]
chart.Correlation(cor(NBA_Subset))
```
Based on the correlation plot,

```{r}
# model=lmer()
```

